Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-29 paper.qmd
writing AER report section

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader needs also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of the extent of this problem come from both empirical and computational analyses. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016]. An attempt at reproducing the computational results of 204 papers in *Science* resulted in success for reproduce the findings for 26% [@stoddenEmpiricalAnalysisJournal2018].

>>>>

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader needs also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of scientific reproducibility in several fields confirm the extent of this problem. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].

Similarly bleak results come from measurements specifically of the reproducibility of computational analyses of scientific studies. An attempt at reproducing the computational results of 204 papers in *Science* resulted in success in reproducing the findings for 26% [@stoddenEmpiricalAnalysisJournal2018]. The computational results of two out of 41 geoscience papers could be fully reproduced on the first attempt [@konkolComputationalReproducibilityGeoscientific2019]. In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024]. Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024]. These efforts confirm that the reproducibility of published research is widely recognised as a cornerstone of rigorous science, and work on evaluating how successful a research community is at generating reproducible results has become a distinctive and important meta-research activity in many fields.

How does archaeology compare to these other fields in terms of reproducibility? Empirical reproducibility has long been valued in field archaeology. Throughout the history of archaeology, well-known sites have been repeatedly revisited to test old hypotheses with new evidence or methods, for example, Olduvai Gorge (Tanzania), Cahokia (USA), Çatalhöyük (Turkey), and Madjedbebe (Australia). Similarly among many experimental archaeologists, empirical reproducibility is a key concern, for example in lithic use-wear identification [@hayesLearningBlindTests2017] and the measurement of lithics [@pargeterReplicabilityLithicAnalysis2023]. To what extent does this concern extend to computational reproducibility among archaeologists?

## Investigating computational reproducibility in archaeology

In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to investigate computational reproducibility in archaeology. In January 2024 I accepted the position of 'Associate Editor for Reproducibility' and conducted peer reviews of submissions that mentioned programming languages such as R or Python in the methods sections.

it is hard to assess whether reported findings are credible.

While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the *American Economic Review*, *American Economic Journal*, *Econometrica*, *Review of Economic Studies*, *Economic Journal*, *Journal of the American Statistical Association*, *American Journal of Political Science*, and the *Journal of Politics*.

Computationally reproducible research can still be plagued by various types of errors, ranging from honest mistakes (e.g., typos in the code) to outright fraud (e.g., data fabrication). A practical hurdle in ascertaining computational reproducibility for the published literature is that the verification is resource-intensive
